---
title: "Homework #5"
author: "Jesse Grigolite"
date: 6/6/23
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: yeti
execute: 
  message: false
  warning: false
---

# Introduction:

*Sarrecenia*, commonly known as the pitcher plant, is a carnivorous plant typical of well-lit and nutrient poor ecosystems (Ellison and Gotelli, 2002).The plant is reliant on prey in order to receive their essential nutrients of Nitrogen and Phosphorous, while still photosynthesizing to harness energy (Wakefield et. al., 2005). For these reasons, *Sarrencia* are equipped with both pitchers, leaves specialized for capturing anthropods and digestion, and phyllodes, leaves specialized in photosynthesis (Ellison and Gotelli, 2002). These plants are important model systems for a variety of reasons. Firstly, they are perennial plants and can live for decades, allowing ecologists to understand how past environmental conditions affect present populations (Ellison and Gotelli, 2002). Secondly, pitcher plants are predominantly located in nutrient deficient areas, therefore addition of nutrients such as nitrogen can have an effect on their physiology and possibly be used an an indicator for nitrogen levels (Ellison and Gotelli, 2002). Finally, *Sarrencia* live in bogs, which are ecosystems with high biodiversity, making the plant an integral part of the health of this ecosystem. In this analysis we will be addressing the question of what *Sarrencia* characteristics best predict total biomass. By understanding the specific morphological, physiological, and taxonomic characteristics that promote pitcher plant biomass, we can understand how to better manage and conserve this species, whether pitcher plants can be nutrient indicators, and what past environmental conditions may have been like for a specific pitcher plant population. We will be testing the null hypothesis that physiological characteristics do not affect pitcher plant biomass, with the alternative hypothesis that at least one physiological characteristic predicts pitcher plant biomass.

# Methods

To conduct this study,ten different species of pitcher plants were used. Two plants of each species, twenty plants in total, were subjected to six different levels of feeding, for a total sample size of 120 plants. All plants were fed their appropriate amount once per week for a total of seven weeks. A total of 32 measurements of physiological characteristics were taken for each plant and included in a data set that will be analyzed in this report (Ellison and Fransworth, 2021). Next, the data set was imported and cleaned up, by renaming columns with easy to read labels and only keeping the morphological characteristics/columns we are interested in. The columns that were selected were total biomass, species, feed level, specific leaf area, chlorophyll, mass based light saturated photosynthetic rate of the youngest leaf, as well as the number of phyllodes and total number of pitchers and phyllodes of each plant.

```{r}
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(performance)
library(naniar) # or equivalent
library(flextable) # or equivalent
library(car)
library(broom)
# would be nice to have
library(corrplot)
library(AICcmodavg)
library(GGally)
library(MuMIn)
```

```{r, reading-data}
plant <- read_csv(here("data", "knb-lter-hfr.109.18", "hf109-01-sarracenia.csv")) %>% 
  #to make colunm names cleaner
  clean_names() %>% 
  #selecting the list of columns to use/keep
  select(totmass, species, feedlevel, sla, chlorophyll, amass, num_lvs, num_phylls)

```

Based on this new data set, we created a visualization of all missing observations.

```{r, missing-data-visualization}
#visualizing the missing data
gg_miss_var(plant)

```

**Figure 1: Data Visualization** Each column is displayed on the y-axis with the number of missing variables on the x-axis. 

The visualization indicated that missing data was present in the chlorophyll, mass based light saturated photosynthetic rate of the youngest leaf (amass), specific leaf area (sla), number of phyllodes (num_phylls), and the total number of phyllodes and pitchers (num_lvs). As these missing observations could interfere with our analysis, we removed them from the data set by subsetting the original data set.  
```{r, subset-drop-NA}
#creating a subset of the data frame without missing values
plant_subset <- plant %>% 
  drop_na(sla, chlorophyll, amass, num_lvs, num_phylls)
```
To determine the relationship between numerical variables in our new data set, we calculated Pearon's r and visually represented correlation using a correlation plot.
```{r, correlation-plot}
#calculate pearson's r for numerical predictors, not response variables and not categorical variables
#created a subset of only the numerical predictors
plant_cor <-plant_subset %>% 
  select(feedlevel:num_phylls) %>% 
  cor(method = "pearson")
#creating a correlation plot
corrplot(plant_cor,
         #change shape of what's in the cells
         method = "ellipse",
        #adds correlation coefficients, r
          addCoef.col = "black"
         )
```
**Figure 2: Pearson's r Correlation Plot** Each predictor variable labels a row or columns, where these rows and columns intersect describes the correlation between the two predictors. The Pearson's r correlation coefficient for the two predictors is written in the box, the direction the ellipse points indicates the direction of the relationship, and the color indicates the strength of the correlation.

The visualization illustrated the strength and direction of the relationships between each of the predictor variables, which if strong can be an indicator of colinearity between predictor variables. All of the correlation values looked within reason so we continued with our analysis.To further explore the data and include categorical as well as numerical variables, another data visualization was created.
```{r, pairs-plot}
plant_subset %>% 
  #can now include categorical variables, putting the variables on x and y axis
  select(species:num_phylls) %>% 
  ggpairs()

```
**Figure 3: Relationships Between Predictor Variables** Each plot in the grid compares two predictor variables. There is a visualization and correlation value (if numerical) for each predictor pair.

The matrix of plots visually represented the relationship between each predictor variable. The scatter plots showed the two variables plotted against another, the box plots are used for a visualization of the categorical variable of species, and the correlations are given for each numerical pair of predictor variables. Since we cleaned up the data, visually reviewed it, and explored relationships amongst predictor variables, we then began with creating our null and full models to use as baselines for our analysis. 
```{r, null-and-full-model}
#creating models for both null and full
null <- lm(totmass ~ 1, data = plant_subset)
full <- lm(totmass ~ species + feedlevel + sla + chlorophyll + amass + num_lvs + num_phylls, data = plant_subset)
```
The null model is a fit linear model to predict total biomass of pitcher plants with no predictors. The full model is the fit linear model to predict total biomass when all predictor variables are factored in. We then checked the statistical assumptions associated with multiple linear regression the full These ass model. These assumptions include independent data, linearity, homoscedasticity, and normality. We first visually assessed normality and homoscedasticity using diagnostic plots for the full model:
```{r, full-diagnostic}
#running diagnostic plots on the full model
par(mfrow = c(2,2))
plot(full)
#based on how clumped the residuals are int he beginning it is border line for homoscedasticity, moreso heterscedastic
```
**Figure 4: Full Model Diagnostics** The two plots on the left visualize residuals, the top right checks for normality, and the bottom right checks for out liars.

Based on the diagnostic plots for the full model above, we found that the residuals were not randomly distributed and appeared heteroscedastic. We checked the full model statistically for normality using the Shapiro-Wilk Test where the null hypothesis is that the model is normally distributed. We also check the full model for homoscedasticity using the Breusch-Pagan test, where the null hypothesis is that the model is homoscedastic.
```{r}
check_normality(full)
#not normal and heteroscedasticity
check_heteroscedasticity(full)
```
Both the Breusch-Pagan Test and Shapiro-Wilk Test had significant p-values, indicating that the null needed to be rejected in both instances. The full model does not meet the assumptions of a multiple linear regression because it is not normally distributed or homoscedastic. In order to meet these assumptions, we mathematically transformed the null and full models by using a log to increase residual normality.
```{r}
#transforming the full and null models
null_log <- lm(log(totmass) ~ 1, data = plant_subset)
full_log <- lm(log(totmass) ~  species + feedlevel + sla + chlorophyll + amass + num_lvs + num_phylls, data = plant_subset)
```
We then checked that the new full logarithmic model met the assumptions visually and statistically.
```{r}
#plotting diagnostics for the log full model
par(mfrow = c(2,2))
plot(full_log)
```
**Figure :5 Full Logarithmic Model Diagnostics** The two plots on the left visualize residuals, the top right checks for normality, and the bottom right checks for out liars.

The residuals appeared much more randomly distributed about the red line, indicating homoscedasticity. We then checked for normailty and homoscedasticity statistically by using the Breusch-Pagan Test and Shapiro-Wilk Test.
```{r}
#statistically testing the full log model
check_normality(full_log)
check_heteroscedasticity(full_log)
```
Both tests then indicated insignificant p-values, allowing us to retain the null and able to claim normality of residuals and homoscedasticity. We then evaluated multicollinearity in the full log model, which can lead to illogical coefficients and large changes in model output, by calculating the generalized variance inflation factor. 
```{r, calculate-vif}
#show GVIF generalized which is for categorical predictors, transformation to compare GVIF to VIF in the last column
#means if there are predictors that are explaining more variance in your response variable because the predictor variables are interacting
car:: vif(full_log)
```
We determined that multicollinearity is not a possibility in the full log model because all of the variance inflation factors were below 5. Next we explored a second model for total pitcher plant biomass that only has the predictor of species. 
```{r, model-1}
#this model only has one predictor that is species to predict total biomass
model2_log <- lm(log(totmass) ~ species, data = plant_subset)
```
As inherit biology and genetics is determinant for total size of many other animals, species seemed to be an appropriate predictor that could explain total biomass simply. The second model was then checked visually for the assumptions of multiple linear regression.
```{r}
#visually checking assumptions
plot(model2_log)
```
**Figure 6: Second Model Diagnostics** The visualization describes the distribution of residuals and normality of data in the second model which includes species as the predictor for pitcher plant total biomass. 

Based on the visualization we believed the model to be normally distributed and homoscedastic. We then checked the second model statistically using the Breusch-Pagan Test and Shapiro-Wilk Test. 
```{r}
#checking for normality of second model
check_normality(model2_log)
#checking homoscedasticity of second model
check_heteroscedasticity(model2_log)
```
As both tests reported insignificant p-values, the second model which only included species as a predictor was concluded to meet all the assumption of multiple linear regression. Next, a third model was formed to predict pitcher plant total biomass.
```{r}
#creating the third model
model3_log <- lm(log(totmass) ~ species + feedlevel, data = plant_subset)
```

The third model included species and feed level as the predictors for biomass. Although genetics determines a large amount of individual physiology, nutrient availability in the form of feed, can also influence growth. Furthermore, pitcher plants are native to nutrient deficient ecosystems making them reliant on carnivore for essential nutrients, therefore an increase in nutrient availability may allow them the opportunity to increase their biomass. The third model was then checked visually for meeting the assumnptions of multiple linear regression.
```{r}
#running visual diagnostics on the third model
plot(model3_log)
```
**Figure 7: Third Model Diagnostics** The visualization describes the distribution of residuals and normality of data in the third model which includes species and feed level to predict total biomass of pitcher plants.

As the visualization displayed randomly distributed residuals about the red line and normally distributed data based on the Q-Q plot, we assumed the third model to meet the necessary assunmptions for multiple linear regression. However, to ensure these assumptions we tested them statistically using the Breusch-Pagan Test and Shapiro-Wilk Test. 
```{r}
#checking for normality in third model
check_normality(model3_log)
#checking for homoscedasticity in third model
check_heteroscedasticity(model3_log)
```
Both tests resulted in insignificant p-values, so the null was retained which indicated that the third model was homoscedastic and had normally distributed residuals. Finally, a fourth model was constructed that included species, total number of leaves (phyllodes and pitchers), and chlorophyll content.
```{r}
#creating model 4
model4_log <- lm(log(totmass) ~ chlorophyll + num_lvs + species, data = plant_subset)
```
Since pitcher plants harness energy from both carnivory and photosynthesis, it was important to include a model that accounted for the latter. Developing biomass, i.e. growing, requires extensive energy, therefore a plant's photosynthetic capabilities, measured in chlorophyll and leaves, could influence and therefore predict its biomass. The model was then checked visually for meeting the requirements of multiple linear regression.
```{r}
#running visual diagnostics on model 4
plot(model4_log)
```
**Figure 8: Fourth Model Diagnostics** The visualization describes the distribution of residuals and normality of data in the fourth model which includes species, chlorophyll, and total number of leaves to predict total biomass of pitcher plants.

Based on the above plots, we interpreted the fourth model to be homoscedastic and normally distributed. To check the assumptions statistically, we then used the Shapiro-Wilk Test to test for normality, and the Breusch-Pagan Test to test for homoscedasticity.
```{r}
#checking fourth model for normality
check_normality(model4_log)
#checking fourth model for homoscedasticity
check_heteroscedasticity(model4_log)
```
In both tests the null was retained, indicating that model four is both normally distributed and homoscedastic, therefore meeting the requirement of multiple linear regression.








# Works Cited

Ellison, Aaron M., and Nicholas J. Gotelli. "Nitrogen Availability Alters the Expression of Carnivory in the Northern Pitcher Plant, Sarracenia Purpurea            Purpurea." Proceedings of the National Academy of Sciences, vol. 99, no. 7, 19 Mar. 2002, pp. 4409--4412, https://doi.org/10.1073/pnas.022057199.

Gotelli, Nicholas J., and Aaron M. Ellison. "Nitrogen Deposition and Extinction Risk in the Northern Pitcher Plant, Sarracenia Purpurea." Ecology, vol. 83, no. 10, 1 Oct. 2002, pp. 2758--2765, https://doi.org/10.1890/0012-9658(2002)083\[2758:ndaeri\]2.0.co;2.

Wakefield, Amy E., et al. "Prey Addition Alters Nutrient Stoichiometry of the Carnivorous Plant Sarracenia Purpurea." Ecology, vol. 86, no. 7, 1 July 2005, pp. 1737--1743, https://doi.org/10.1890/04-1673.
